---
title: "HW2: Surafel Geleta \nssg2775"
output:
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Calling packages
library(tidyverse)
library(kableExtra)

# Importing datasets

profs <- read.csv("C:\\Users\\gelet\\OneDrive\\Documents\\SDS315\\HW2\\profs.csv")
bike_share <- read.csv("C:\\Users\\gelet\\OneDrive\\Documents\\SDS315\\HW2\\bikeshare.csv")
capmetro_UT <- read.csv("C:\\Users\\gelet\\OneDrive\\Documents\\SDS315\\HW2\\capmetro_UT.csv")
billboard <- read.csv("C:\\Users\\gelet\\OneDrive\\Documents\\SDS315\\HW2\\billboard.csv")

```

# Problem 1: Beauty, or not, in the classroom

## Part A

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(profs, aes(x = eval)) +
  geom_histogram(fill = "#0072B2", color = 'black') +
  labs(x = "Instructor Average Teaching Evaluation Score",
       y = "Number of Courses", 
       title = "Distribution of Sampled UT Teaching Evaluation Scores")
```

Instructors at UT Austin in the sample have a median average teaching evaluations score of 4, indicating that most of the sampled instructors are rated generally positively. The distribution of scores is left-skewed, and the IQR of average scores is 0.8, showing that 50% of instructors fall within 0.8 points of the median score.

## Part B

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(profs, aes(x = factor(native), y = eval)) +
  geom_boxplot(fill = "#0072B2", color = 'black') +
  labs(x = "Is the Instructor a Native English Speaker?",
       y = "Instructor Average Teaching Evaluation Score",
       title = "Distribution of Sampled UT Teaching Scores by Instructor English")
```

Sampled UT instructors who are native English speakers have a median average evaluation score 0.4 points higher than non-native English speaking instructors.

## Part C

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(profs, aes(x = eval)) +
  geom_histogram(fill = "#0072B2", color = 'black') +
  labs(x = "Instructor Average Teaching Evaluation Score",
       y = "Number of Courses",
       title = "Sampled UT Teacher Evaluation Score by Instructor Gender") +
  facet_wrap(vars(gender), nrow = 2)
```

Sampled male instructors had a median average evaluation score 0.25 points higher than sampled female instructors. The distributions for both genders are left-skewed, and male instructors have a slightly wider range of evaluation scores than female instructors, with an IQR 0.1 points lower.

## Part D

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(profs, aes(x = beauty, y = eval)) +
  geom_point(size = 2) +
  labs(x = "Instructor Average Attractiveness Rating",
       y = "Instructor Average Teaching Evaluation Score",
       title = "Relationship between Instructor Evaluation Score and Beauty")
```
There is a very weak, positive relationship between sampled instructor attractiveness and teaching evaluation score, as indicated by the correlation coefficient value of `r cor(profs$beauty, profs$eval)`.

# Problem 2: Bike Sharing

## Plot A

```{r echo=FALSE, message=FALSE, warning=FALSE}
bike_share %>% 
  group_by(hr) %>% 
  summarize(average = mean(total)) %>% 
  ggplot(aes(x = hr, y = average)) + 
  geom_line() +
  labs(x = "Hour of Day (0 = midnight)",
       y = "Average Number of Bike Rentals",
       title = "D.C. Capital Bikeshare Rentals Across Hours, 2011-12")
```

## Plot B

```{r echo=FALSE, message=FALSE, warning=FALSE}
bike_share %>% 
  group_by(hr, workingday) %>% 
  summarize(average = mean(total),
            workingday = workingday) %>%
  # Recoding workingday values for better comprehension
  mutate(workingday = recode(workingday,
                             "0" = "Non-Working Day",
                             "1" = "Working Day")) %>% 
  ggplot(aes(x = hr, y = average)) + 
  geom_line() +
  labs(x = "Hour of Day (0 = midnight)",
       y = "Average Number of Bike Rentals",
       title = "D.C. Capital Bikeshare Rentals Across Hours by Day, 2011-12") +
  # Faceting by day
  facet_wrap(vars(workingday))

bike_share %>% 
  group_by(hr) %>% 
  summarize(average = mean(total),
            workingday = workingday)
```


## Plot C

```{r echo=FALSE, message=FALSE, warning=FALSE}
bike_share %>% 
  filter(hr == 9) %>% 
  group_by(workingday, weathersit) %>% 
  summarize(average = mean(total),
            weathersit = weathersit) %>%
  # Removes duplicated rows
  distinct() %>% 
  mutate(workingday = recode(workingday,
                             "0" = "Non-Working Day",
                             "1" = "Working Day")) %>% 
  ggplot(aes(x = weathersit, y = average)) + 
  geom_bar(fill = "#0072B2", color = 'black', stat = "identity") +
  labs(x = "Weather Situation Code",
       y = "Average Number of Bike Rentals",
       title = "D.C. Capital Bikeshare Rentals by Weather and Day, 2011-12") +
  facet_wrap(vars(workingday))

# Table for reference to values

bike_share %>% 
  filter(hr == 9) %>% 
  group_by(hr, workingday, weathersit) %>% 
  summarize(average = mean(total),
            weathersit = weathersit) %>% 
  distinct()
```

# Problem 3: Capital Metro UT Ridership

## Plot 1

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Recoding categorical variables in sensible rather than alphabetical order
capmetro_UT = mutate(capmetro_UT,
day_of_week = factor(day_of_week,
levels=c("Mon", "Tue", "Wed","Thu", "Fri", "Sat", "Sun")),
month = factor(month,
levels=c("Sep", "Oct","Nov")))

# Wrangling and Plotting
capmetro_UT %>% 
  group_by(hour_of_day, month, day_of_week) %>% 
  summarise(average = mean(boarding)) %>% 
  ggplot(aes(x = hour_of_day, y = average, group = month, color = month)) +
  geom_line() +
  labs(x = "Hour of Day (6 = 6AM)",
       y = "Average Number of Boarders",
       title = "CapMetro UT Austin Ridership Across Hours, Sep-Nov 2018") + 
  facet_wrap(vars(day_of_week))

# Table for reference to values

capmetro_UT %>% 
  group_by(hour_of_day, month, day_of_week) %>% 
  summarise(average = mean(boarding))

```
This graphic displays CapMetro’s bus ridership trends across stops around UT Austin between September and November of 2018. The y-axis shows the average, or mean, number of boarders for a given hour (Hours start at 6AM, and end at 9PM), and the x-axis shows bus operating hours within each day. The graphic is faceted, displaying 7 plots corresponding to data from each day of the week, with each plot having three color-coded lines corresponding to a month. Saturday and Sunday ridership is much lower than on weekdays; moreover, weekend ridership does not seem to follow the same patterns as weekday ridership. For instance, on Saturday average ridership peaks at 5PM in October, with an average of 26 riders within that hour; that same hour and month on Friday sees an average of around 127 riders within the hour. Weekday ridership generally follows a similar pattern, with boarding starting low at 6AM, followed by an increase in boardings until reaching peak ridership, typically between 3-5PM. From there, boardings drop steeply into the night. Peak hours vary slightly depending on the day, as well as the month; on Mondays, September boardings peak at 5PM, and October and November at 4PM. Tuesday September boardings peak at 4PM, October at 5PM, and November at 4PM. September and November Thursdays experience peak boarding during 3PM. Saturday and Sunday stray a little from this trend; in particular, peak boardings during November Saturdays occurred at 11AM, while peaks on September and October Sundays occurring at 7PM and 1PM respectively. September boardings are noticebly than October and November boardings on Mondays; this may be explained by Labor Day, which falls every 1st Monday of September, contributing to off-days from school and work. Similarly, November boardings during Wendesday-Friday are generally lower than the other months; this is likely due to Thanksgiving, which falls on the fourth Thursday of November. Wednesday may experience lower ridership due to people staying home or travelling to prepare for Thanksgiving day on Thursday, and people may rest and recover on Friday.

## Plot 2

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(capmetro_UT, aes(x = temperature, y = boarding, color = weekend)) +
  geom_point() +
  labs(x = "Temperature (degrees Fahrenheit)",
       y = "Number of Boarders",
       title = "CapMetro UT Boarding and Temp. Trends, Sep-Nov 2018") +
  facet_wrap(vars(hour_of_day))

# For loop calculating correlation coefficient
for(i in 6:21){
  print(cor(capmetro_UT$temperature[capmetro_UT$hour_of_day == i], capmetro_UT$boarding[capmetro_UT$hour_of_day == i]))
}

```
This graphic displays the relationship between temperature and total boarding in CapMetro’s UT Austin routes, faceted to include a plot of the relationship for each operating hour of the day. Temperatures increase as the day approaches mid-afternoon (3-5PM), seen with how the points generally shift right between 6AM and 5PM; temperatures begin to decline after 5PM. In addition, the plots indicate that weekend boarding counts are significantly lower than weekday boardings across all available hours. When holding the hour of the day and day type constant, there is little evidence of an association between temperature and the number of boarders. The Pearson correlation coefficient values across all hours are incredibly low, varying between -0.0682 to 0.0683, which indicate no linear relationship.

# Problem 4: Wrangling the Billboard Top 100

## Part A

```{r echo=FALSE, message=FALSE, warning=FALSE}
topten<- billboard %>% 
  group_by(performer, song) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  slice_max(count, n = 10)

# Creating Table
knitr::kable(topten,
             col.names = c("Performer", "Song", "Count"),
             caption = "Top 10 Most Popular Songs By Count of Weeks Spent on Billboard Top 100") %>% 
  # Table aesthetic
  kable_classic(full_width = F, html_font = "Cambria")
```


## Part B

```{r echo=FALSE, message=FALSE, warning=FALSE}
billboard %>%
  group_by(year) %>% 
  summarize(uniques = length(unique(song)),
            year = year) %>% 
  ggplot(aes(x = year, y = uniques)) +
  geom_line() +
  labs(x = "Year",
       y = "Number of Unique Songs Appearing on Billboard Top 100",
       title = "'Music Diversity' in the Billboard Top 100, 1959-2020")
```


## Part C

```{r echo=FALSE, message=FALSE, warning=FALSE}
#billboard %>% 
  
```

